\documentclass[lotsofwhite]{patmorin}
\input{pat.tex}

\title{\MakeUppercase{Diameter and Smallest Enclosing Circle Peeling Depth}}
\author{Prosenjit Bose \and Pat Morin}
\date{}

\begin{document}
\maketitle
\begin{abstract}
We introduce two new notions of data depth based on a generalization
of the fact that the median is obtained by repeatedly removing the
minimum and maximum of a set until only one or two elements remain.
\end{abstract}

\section{Introduction}

Let $S$ be a set of $n$ points in $\mathbb{R}^d$.  The \emph{diameter} 
of a point set $d(S)$ is defined as
\[
    d(S)=\max\left\{\|xy\|:x,y\in S\right\} \enspace .
\]  
A pair of points $x$ and $y$ such that
$\|xy\|=D(S)$ is called a \emph{diametrical pair}. Let $D_0=S$ and
let $D_i$ be obtained from $D_{i-1}$ by removing a diametrical pair.
The \emph{diameter peeling depth} of $x$ with respect to $S$ is the
maximum value $i$ such that $x\in D_i$.

The smallest enclosing ball $B(S)$ is the smallest closed ball that
contains all points of $S$.  The radius of $B(S)$, denoted $r(S)$ is
called the \emph{radius} of $S$.  Note that, in general, $B(S)$ has
anywhere between $2$ and $d+1$ points of $S$ on its boundary.  Let
$B_0=S$ and let $B_i$ be obtained from $B_{i-1}$ by removing all
points on the boundary of $B(B_{i-1})$.  The \emph{smallest enclosing
ball peeling depth} of $x$ with respect to $S$ is the maximum value
$i$ such that $x\in B_{i}$.


\section{Diameter-Peeling Depth}

\begin{thm}
Let $S$ be a set of $n$ points in $\mathbb{R}^d$ and let $X\subset S$ and
$Y$ each be of size at most $n/2-1$.  Then the diameter-peeling center
of $S\cup Y\setminus X$ is contained in the convex hull of $S$.
\end{thm}


\begin{thm}
Let $S$ be a set of $n$ points in $\mathbb{R}^d$.  The
diameter-peeling depth of every point in $S$ can be computed in
$O(dn^2\log n)$ time.
\end{thm}

\begin{thm}
Let $S$ be a set of $n$ points in $\mathbb{R}^3$.  The
diameter-peeling depth of every
point in $S$ can be computed in $O(n\log^{O(1)} n)$ time.
\end{thm}

\begin{thm}
Let $S$ be a set of $n$ points in $\mathbb{R}^2$.  The
diameter-peeling depth of every point in $S$ can be computed in
$O(n\log n)$ time.
\end{thm}


\section{Smallest Enclosing Circle-Peeling Depth}

\begin{thm}
Let $S$ be a set of $n$ points in $\mathbb{R}^d$ and let $X\subset S$ and
$Y$ each be of size at most $n/3-1$.  Then the smallest enclosing
ball-peeling center of $S\cup Y\setminus X$ is contained in the convex
hull of $S$.
\end{thm}

\begin{thm}
Let $S$ be a set of $n$ points in $\mathbb{R}^d$.  The
smallest enclosing ball-peeling depth of every point in $S$ 
can be computed in $O(d!n^2)$ time.
\end{thm}

\begin{thm}
Let $S$ be a set of $n$ points in $\mathbb{R}^2$.  The
smallest enclosing ball-peeling depth of every point in $S$ 
can be computed in $O(n^{3/2}\log^{O(1)} n)$ time.
\end{thm}



\section{Conclusions}
\end{document}
