\documentclass{article}

\title{Guest Editors' Introduction}
\author{Prosenjit Bose \and Pat Morin}
\date{}

\setlength{\parskip}{1ex}

\begin{document}
\maketitle

This special issue of Theory of Computing Systems is made up of
selected papers from the 13th International Symposium on Algorithms
and Computation (ISAAC~2002) that took place in Vancouver in November
2002.  Of the nearly 160 papers submitted to the symposium, 54 were
accepted and 6 of those were selected to appear in this special issue.

By coincidence, all of the papers in this special issue are related to
the ever expanding field of (geometric and abstract) graph theory and
all have exactly two authors.  \mbox{Biedl} and \mbox{Wilkinson}
present results on finding large bounded-degree independent sets in
planar graphs.  \mbox{Amb\"uhl} and \mbox{Wagner} show new results on
the hardness of the Clique Problem in intersection graphs of geometric
objects.  \mbox{Buchheim} and \mbox{Hong} give results on finding
symmetric drawings of graphs with few crossings.  \mbox{Erten} and
\mbox{Kobourov} show how to simultaneously embed a planar graph and
its dual on a small integer grid.  \mbox{Avidor} and \mbox{Zwick} give
improved bounds for the approximation ratios of MIN~2-SAT and
MIN~3-SAT.  Finally, \mbox{Jurdzi\'nski} and \mbox{Stachowiak} give
algorithms for the wakeup problem in wireless networks.

Writing high-quality papers is a time-consuming task and so is
thorough refereeing.  We are grateful to the authors in this special
issue for preparing such excellent papers and to the anonymous
referees for all their hard work.  We hope the reader enjoys this
special issue as much as we enjoyed preparing it.

\noindent
Prosenjit Bose and Pat Morin \\
Ottawa, November 2004

\end{document}
