\documentclass{patmorin}
\usepackage{pat,amsopn}

\DeclareMathOperator{\height}{height}


\title{Notes on Comparison-Based Sorting Lower Bounds}
\author{\ }


\begin{document}
\maketitle

\begin{abstract}
These are some notes from working with John on comparison-based sorting.
\end{abstract}

\section{Introduction}

We want to prove the following theorem (which is not true):

\begin{thm}\thmlabel{lower-bound}
  For any deterministic comparison-based sorting algorithm, $\mathcal{A}$,
  there exists a permutation of $\{1,\ldots,n\}$ on which $\mathcal{A}$
  performs at least $\log(n!)+n/9-O(\log n)$ comparisons.
\end{thm}

\section{Preliminaries}

A \emph{comparison tree}, $T$, for sorting $n$ elements is a binary tree.
Each internal node, $u$, of $T$ is labelled with an ordered pair of
indices $(i_u,j_u)$.  Each leaf, $\ell$, of $T$ is labelled with a
permutation, $\pi_\ell$, of $\{1,\ldots,n\}$.

Every sequence, $a=a_1,\ldots,a_n$, of numbers defines a root-to-leaf
path, $T(a)$, in a comparison tree, $T$, as follows:  The first node
in $T(a)$ is the root.  When situated at an internal node $u$, the path
$T(a)$ continues to $u$'s left child if $i_u < j_u$ and continues to
$u$'s right child otherwise.

We say that the comparison tree $T$ \emph{sorts a set of sequences}, $A$,
if, for every sequence $a_1,\ldots,a_n\in A$, $T(a_1,\ldots,a_n)$ ends at
a leaf, $\ell$, that is labelled with a permutation $\pi_\ell$ such that:
\[
    a_{\pi_\ell(1)} \le a_{\pi_\ell(2)} 
        \le \cdots \le a_{\pi_\ell(n)} \enspace .
\]
We say, simply, that $T$ \emph{sorts} if it sorts the set $\R^n$.  It is
not hard to verify that, for $T$ to sort, it is sufficient for $T$ to sort
the set, $\Pi_n$, of all permutations of the integers $\{1,\ldots,n\}$. We call the elements of $\Pi_n$ \emph{inputs}.

Note that, for any node, $u$, in a comparison tree, there exists a maximal
set of inputs, $A(u)$, such that $T(a)$ contains $u$ for all $a\in A$.
That is, every permutation in $A(u)$ leads to $u$.  In this paper, we
are interested in comparison trees of minimum height. We can therefore
assume that our comparison trees have the following properties:

\begin{prp}[No unreachable nodes]\prplabel{can-reach}\prplabel{first}
  For any node $u$ in $T$, $A(u)\neq\emptyset$.
\end{prp}

\begin{prp}[No useless comparisons]
  Each internal node $u$ of $T$ has at least 2 children.
\end{prp}

\begin{prp}[Subtree optimality]\prplabel{last}
  For any node, $u$ of $T$, the subtree rooted at $u$ is a comparison tree
  of minimum height that sorts $A(u)$. 
\end{prp}

It is well known that, for any deterministic comparison-based sorting
algorithm, $\mathcal{A}$, the actions of $\mathcal{A}$ can be modelled by
a comparison tree, $T_\mathcal{A}$.  Thus, to prove \thmref{lower-bound},
it suffices to prove the following equivalent lemma:

\begin{lem}\lemlabel{lower-bound}
  Every comparison tree that sorts $\Pi_n$ has height at least
  $\log(n!)+n/9-O(\log n)$.
\end{lem}

\section{The Lower Bound}

The sketch of our lower-bound argument is as follows:  We first show
that, for any comparison tree, $T$, that sorts, there exists an input
$a=a_1,\ldots,a_n$ such that $T(a)$ has length $\log(n!)-O(\log n)$ and
that there are $n/9-O(\log n)$ comparisons performed along the path $T(a)$
that are \emph{lucky} (in a formal sense, defined below).  We then use
the existence of these efficient comparisons to derive another comparison
tree $T'$ and another permutation $a'$ with the following two properties,
which establish \lemref{lower-bound}:
\begin{enumerate}
  \item The height of $T'$ does not exceed the height of $T$.
  \item $T'(a')$ has length at least $\log(n!)+n/9-O(\log n)$.
\end{enumerate}
In particular, these two properties imply that
\[
   \height(T) \ge \height(T') \ge \log(n!)+n/9-O(\log n) \enspace .
\]
(The first inequality is implied by the first property and the second
inequality is implied by the second property.)  We emphasize that the
tree $T'$ does not necessarily sort any sequence other than $a'$.

\subsection{Lucky and Unlucky Triples}

Our proof relies on grouping elements of the input $a$ into $n/3$ triples.
For the remainder of the paper we will assume, only for simplicity,
that $n\equiv 0\pmod{3}$. 

We need a notation to refer to the position of the item $i$ in an input
$a=a_1,\ldots,a_n$.  For this purpose, we define the \emph{inverse}
$\bar{a}=\bar{a}_1,\ldots,\bar{a}_n$ so that $\bar{a}_i$ is the unique
index $j$ such that $a_j=i$.

We group the indices of $a$ into $n/3$ triples
(sets of size 3), $t_1,\ldots,t_{n/3}$, where
$t_i=\{\bar{a}_{3i-2},\bar{a}_{3i-1},\bar{a}_{3i}\}$.  That is, each
$t_i$ contains the indices of 3 consecutive values in $\{1,\ldots,n\}$.
Triples are useful, because $T$ can gain no information about the relative
ordering of the elements in a triple except by comparing two elements,
both of which are in the triple.

If $T$ sorts then, for any input $a$ and any triple $t_i$ of $a$,
$T(a)$ must contain at least 2 nodes $u$ such that $i_u$ and $j_u$
are both in $t_i$.  In some cases, $T(a)$, contains three such nodes.
In the former case, we say that $T$ \emph{got lucky} for the triple $t_i$
and we say that $T$ was \emph{unlucky} (for $t_i$) otherwise.  In the
case where $T$ is lucky for $t_i$, we call the second node on $T(a)$
that compares two indices in $t_i$ a \emph{lucky node}.

\begin{lem}\lemlabel{triples}
  Let $a_1,\ldots,a_n$ be a random permutation of $\{1,\ldots,n\}$.
  Then the expected number of lucky nodes in $T(a)$ is $n/9$ and the
  expected number of unlucky nodes in $T(a)$ is $2n/9$.
\end{lem}

\begin{proof}
  For a triple $t_i=\{i_1,i_2,i_3\}$, let $x=a_{i_1}$, $y=a_{i_2}$,
  and $z=a_{i_3}$.  Assume, without loss of generality that the
  first comparison within $t_i$ determines that $x<y$ and that
  the second comparison is between $y$ and $z$ (other cases are
  symmetric).  Conditioned only on the assumption, $x<y$, each of
  the tree possibilities for the location of $z$ is equally likely.
  In particular, the relative ordering
  \[
     x < y < z
  \]
  occurs with probability $1/3$.  This is the only case in which the
  total ordering of $x$, $y$, and $z$ can be determined with only
  two comparisons.  In the other cases (with probability $2/3$), an
  additional comparison is required to distinguish between
  \[
     x < z < y
  \]
  and 
  \[
     z < x < y \enspace .
  \]
  Thus with probability $1/3$ $T$ is lucky for $t_i$ and with probability
  $2/3$ $T$ is unlucky.  The lemma then follows from linearity of
  expectation.
\end{proof}

\begin{lem}\lemlabel{permutation}
  For any decision tree $T$, the exists an input $a=a_1,\ldots,a_n$
  such that the root-to-leaf path defined by $a$ has length at least
  $\log(n!)-O(\log n)$ and the number of triples in $a$ on which $T$
  gets lucky is at least $n/9-O(\log n)$.
\end{lem}

\begin{proof}
  Let $a$ be a random input and fix an integer constant, $c\ge $, to be
  defined later.  Let $\mathcal{E}_1$ be the event
  ``the length of $T(a)$ is at least $\log(n!)-c$.''  Then,
  \[
     \Pr\{\mathcal{E}_1\} \ge 1-2^{-c} \enspace ,
  \]
  since any binary tree contains at most $k/2^c$ leaves of depth at most
  $\log(k)-c$.)
  
  Let $\mathcal{E}_2$ be the event ``$T$ is lucky on at least
  $(1/3-2\epsilon)n/3$ triples of $a$.'' Then, by Markov's Inequality and
  \lemref{triples}
  \[
     \Pr\{\mathcal{E}_2\} \ge \epsilon \enspace .
  \]
  
  Therefore, if $\epsilon > 2^{-c}$, then this implies that
  \[
      \Pr\{\mbox{$\mathcal{E}_1$ and $\mathcal{E}_2$}\} > 0 \enspace ,
  \]
  and hence there exists at least one input $a'$ for which both
  conditions hold.  We complete the proof by taking $\epsilon=\log n/ n$
  and $c=\log(1/\epsilon)+1$.
\end{proof}

\subsection{Finding a Long Path in $T$}

Next, we show how the permutation, $a$, guaranteed by
\lemref{permutation}, can be used to show that $T$ has height
$\log(n!)+n/9-O(\log n)$.  Recall that $T$ is a decision tree that
sorts and that satisfies Properties~\ref{prp:first}--\ref{prp:last}.
This means that $T$ contains no unreachable nodes, never performs a
comparison that can be deduced from previous comparisons, and every
subtree rooted at any node $u$ or $T$ has minimum height among all
comparison trees that sort $A(u)$.

Sticking to our plan, we now construct the tree $T'$ whose height
does not exceed that of $T$ and that contains a root to leaf path of
length $\log(n!)+n/9-O(\log n)$.  We emphasize, again, that $T'$ does
not necessarily sort any inputs with the exception of one input $a'$
that we derive from $a$ during the construction of $T'$.  The input $a'$
will have the same triple structure as $a$; The $i$th triple of $a$ is the
same as the $i$th triple in $a'$. However, the ordering of the elements
compared at every node $u$ at which $T$ gets lucky will be reversed.
That is, if $T$ gets lucky at node $u$ then $a'$ swaps the values of
$a_{i_u}$ and $a_{j_u}$ so that $a'_{i_u}=a_{j_u}$ and $a'_{i_u}=a_{j_u}$.

To construct $T'$ we copy the path $T(a)$ up to and including the first
vertex $u$ at which $T$ gets lucky for some triple, $t_i$.  At this
point, $T(a)$ continues to $u$'s (say) left child, $w$, since (say) $i_u
< j_u$.  The path $T'(a')$ will now continue to the right child of $w$ since
$a'$ swaps the values $a_{i_u}$ and $a_{j_u}$.

Note that $A(w)$ contains only inputs in which the relative order of the
three elements in $t_i$ is fixed.  On the other hand, for $u$'s right
child, $w'$, $A(w')$ contains inputs in which the three elements of $t_i$
occur in two different orders.

This is where the argument breaks: Our plan was to have $T'$ do a
comparison at $w'$ and then copy the subtree rooted at $w$.  This was
supposed to work, since after performing this extra comparison, the
information known about the permutation is exactly the same as it was
in $w$.  Unfortunately, this doesn't work.  The height of the subtree at
$w'$ may be less than or equal to the height of the subtree rooted at $w$.
In which case, performing the extra comparison yields a tree $T'$ whose
height is greater than that of $T$.
\end{document}
